Litrepl: Literate Paper Processor Promoting Transparency More Than Reproducibility

### 2025-01-05 paper/paper.tex
Update paper

The concept of \textit{Literate Programming} was formulated by Donald Knuth,
suggesting a shift in focus from writing code to explaining to human beings what
we want a computer to do (\textcite{Knuth1984lp}). This approach is embodied in
the WEB system and its descendant family of tools, which translate combined code
and narrative documents into human-readable documentation and machine-executable
code. Over time, the concept has evolved, showing a trend towards simplification
(\textcite{Ramsey1994lps}).

@@

A significant milestone in this field was the IPython interpreter
(\textcite{Perez2007IPython}), which later evolved into the Jupyter Project. The
creators introduced a new document format called the Jupyter Notebook
(\textcite{Kluyver2016jupnb}), characterized by a series of logical sections of
various types, including text and code, which could directly interact with
programming language interpreters. This interactive communication, akin to REPL
style programming, allows the creation of well-structured documents suitable for
presentations and sharing. The concept underpinning these developments is termed
\textit{Literate Computing}\cite{Perez2015blog}, which includes goals of
spanning a wide audience range, boosting reproducibility, and fostering
collaboration. To achieve these objectives, several technical decisions were
made, notably the introduction of bidirectional communication between the
computational core, known as the Jupyter Kernel, and the Notebook web-based
renderer, along with another layer of client-server communication between the
web-server and the user’s web browser.

While we recognize the importance of all goals within the Literate Computing
framework, we argue that the goal of reproducibility is paramount. Successfully
addressing reproducibility would alone suffice to enhance communication among
time-separated and space-separated researchers and significantly expand the
audience. However, as it became clear (\textcite{Dolstra2010}), this challenge
extends beyond the scope of a single human-computer interaction system, and even
beyond the typical boundaries of software distribution management for a
particular programming language. A comprehensive solution to the software
deployment problem should operate at the entire operating system level.

Following \textcite{Vallet2022}, we suggest changing the focus of human-computer
interaction towards simplicity and transparency. We saw an opportunity to
implement a tool that would offer REPL-style editing, be compatible with
existing code editors and text formats, thus maintaining familiar editing
practices, and have a significantly smaller codebase.

@@

In Litrepl, these details are hardcoded for several prominent interpreter
families, which we refer to as \textit{interpreter classes}. At the time of
writing, Litrepl supports three such classes: \textbf{python}, \textbf{sh}, and
\textbf{ai}. Using command line arguments, users can configure how to map code
section labels to the correct class and specify which interpreter command to
execute to start a session for each class.

@@

\begin{figure*}[hbt!]
  \centering
  \begin{minipage}{\textwidth}
  %lresult
  \begin{verbatim}
  \begin{result}
  ... some output ...
  [LR:/tmp/litrepl/python/partial_7b81e1e.txt]
  \end{result}
  \end{verbatim}
  %lnoresult
  \end{minipage}
  \caption{Partial result section with a continuation tag.}
  \label{fig:partial-result}
\end{figure*}

@@

Litrepl provides \textbf{start}, \textbf{stop}, and \textbf{restart} commands to
control background sessions. The \textbf{interrupt} command sends an
interruption signal to the interpreter. Finally, the \textbf{repl} command
establishes direct communication with the interpreter, allowing manual
inspection of its state.
%lnoresult

@@

Evaluation results are encoded back into the result sections, and the entire
document is printed. At this stage, certain conditions can be optionally
checked. First, setting \verb|--pending-exitcode| to a non-zero value instructs
Litrepl to report an error if a section takes longer than the timeout to
evaluate. Second, setting \verb|--exception-exitcode| directs Litrepl to detect
Python exceptions. Lastly, \verb|--irreproducible-exitcode| triggers an error if
the evaluation result doesn't match the text initially present in the result
section.

@@

% lresult
The technical decision to abstract interpreters using text streams comes with
both advantages and disadvantages. A key advantage is simplicity. However, there
are notable negative aspects. Firstly, there is no parallel evaluation at the
communication level, meaning the interpreter is locked until it completes the
evaluation of one snippet before proceeding to the next. Secondly, the
transferable data type is restricted to text-only streams.

>>>>

The concept of \textit{Literate Programming} was formulated by Donald Knuth,
suggesting a shift in focus from writing code to explaining to human beings what
we want a computer to do. This approach is embodied in the WEB
system\cite{Knuth1984lp} and its descendant family of tools, whose name refers
to a text document format containing the "network" of code sections interleaved
with text.

The system could both render such text into human-readable documentation and
compile machine-executable code. Over time, this concept has evolved, showing a
trend towards simplification (\textcite{Ramsey1994lps}).

@@

A significant milestone in this field was the IPython interpreter
(\textcite{Perez2007IPython}), which later evolved into the Jupyter Project. Its
creators introduced a new document format called the Jupyter Notebook
(\textcite{Kluyver2016jupnb}), characterized by a series of logical sections of
various types, including text and code, which could directly interact with
programming language interpreters. This interactive communication, akin to REPL
style programming, allows the creation of well-structured documents suitable for
presentations and sharing. The concept underpinning these developments is termed
\textit{Literate Computing}\cite{Perez2015blog}, which includes goals of
spanning a wide audience range, boosting reproducibility, and fostering
collaboration. To achieve these objectives, several technical decisions were
made, notably the introduction of bidirectional communication between the
computational core, known as the Jupyter Kernel, and the Notebook web-based
renderer, along with another layer of client-server communication between the
web-server and the user’s web browser.

While we recognize the importance of all goals within the Literate Computing
framework, we think that the goal of reproducibility is more important than
others. Addressing it alone would suffice to enhance communication among
time-separated and space-separated researchers and significantly expand the
audience. However, as it became clear (\textcite{Dolstra2010}), this challenge
extends beyond the scope of a single human-computer interaction system, and even
beyond the typical boundaries of software distribution management for a
particular programming language. A comprehensive solution to the software
deployment problem operates at the entire operating system level.

Following \textcite{Vallet2022}, we suggest changing the focus of human-computer
interaction towards simplicity and transparency. We saw an opportunity to
implement a tool that would offer REPL-style editing, be compatible with
existing code editors and text formats, thus maintaining familiar editing
practices, contain no important hidden state variables and have a significantly
smaller codebase.

@@

In Litrepl, these details are hardcoded for several prominent interpreter
families, which we refer to as \textit{interpreter classes}. At the time of
writing, Litrepl supports three such classes: \texttt{python}, \texttt{sh}, and
\texttt{ai}. Using these names in command line arguments, users can configure
how to map code section labels to the correct class and specify which
interpreter command to execute to start a session for each class. For example,
to select a Bourne-Again Shell interpreter as \verb|sh|, we add the
\verb|--sh-interpreter=/usr/bin/bash| argument assuming that this binary is
present in the system.
%lnoresult

@@

\begin{figure*}[hbt!]
  \centering
  \begin{minipage}{\textwidth}
  %lresult
  \begin{verbatim}
  \begin{result}
  ... some output ...
  [LR:/tmp/litrepl/python/partial_7b81e1e.txt]
  \end{result}
  \end{verbatim}
  %lnoresult
  \end{minipage}
  \caption{Example partial result section that ends with a continuation tag,
  where \texttt{LR:} is a distinguishable prefix, and the rest is the filename
  storing a response of the interpreter to the section being evaluated.}
  \label{fig:partial-result}
\end{figure*}

@@

As already mentioned, most of the resources are exposed to user via the
filesystem. The only added hidden state in the entire system resides in the
reader process, in a small finite state machine that handles special ASCII
characters used by some applications for text animation.

Litrepl provides \textbf{start}, \textbf{stop}, \textbf{restart} and
\textbf{status} commands to control background sessions, so, for example

\begin{verbatim}
$ litrepl restart python
\end{verbatim}

stops the Python interpreter if it was running and starts a new instance of it.
The \textbf{interrupt} command sends an interruption signal to the interpreter.
Finally, the \textbf{repl} command establishes direct communication with the
interpreter, allowing manual inspection of its state. The \textbf{help} command
prints the detailed description of each command and the configuration
arguments available.
%lnoresult

@@

Evaluation results are written back into the result sections, and the entire
document is printed. At this stage, certain conditions can be optionally
checked. First, setting \verb|--pending-exitcode| to a non-zero value instructs
Litrepl to report an error if a section takes longer than the timeout to
evaluate. Second, setting \verb|--exception-exitcode| directs Litrepl to detect
Python exceptions. Lastly, \verb|--irreproducible-exitcode| triggers an error if
the evaluation result doesn't match the text initially present in the result
section.

@@

% lresult
The technical decision to abstract interpreters using text streams comes with
both advantages and disadvantages. A key advantage is simplicity. However, there
are notable negative aspects. First, there is no parallel evaluation at the
communication level, meaning the interpreter is locked until it completes the
evaluation of one snippet before proceeding to the next. Second, the
transferable data type is restricted to text-only streams.
